---
title: "P7_Survival_After_Indwelling_Arterial_Catheters_Use"
author: "Dominik Frei"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=10, fig.height=8, fig.fullwidth=TRUE)
```


# Code

## Set up

Libraries:
```{r message=FALSE}
library(tidyverse)
library(Amelia) #missmap()
library(mice) #imputation
```
Language:
```{r}
Sys.setenv(LANG = "en")
```

## Load and Preprocess Data

Load in the data set:
```{r}
data <- read.csv(file = "Data/full_cohort_data.csv",
                 stringsAsFactors = TRUE)
```

Convert all the binary variables into factors:
```{r define factors}
for (i in 1:ncol(data)) {
        if(length(unique(data[,i])) == 2) {
                data[,i] <- as.factor(data[,i])
        }
}
```


Display the data structure as imported:
```{r}
str(data)
```
`gender_num` and `sepsis_flg` should also be factors:
```{r}
data$gender_num <- as.factor(data$gender_num)
data$sepsis_flg <- as.factor(data$sepsis_flg)

data[,c("gender_num", "sepsis_flg")] %>% str()
```
`sepsis_flg` will not be useful in our analysis, since all the values are the same
(it has zero variance). We can therefore drop it from our data:
```{r}
data <- data %>% select(!sepsis_flg)
```

## Exploratory Data Analysis

Lets look at the missing value sof the data:
```{r missmap all}
missmap(data[,colSums(is.na(data)) > 0], main = "Missingness Map: All Patients")
```
The `bmi` variable contains around 400 NA values, which have adjacent indexes.
This is somewhat odd. It does not seem that the data was ordered for missing values
in `bmi`, since there are other NAs, outside of the missing chunk.

Is there anything special about these missing records?

The variable `weight_first` might be somehow related to bmi. Lets plot it versus the rowindex:
```{r}
data %>% ggplot(aes(x = as.numeric(rownames(.)), y = weight_first)) +
        geom_point() +
        theme_classic()
```
There doesnt seem to be a particular order to the weights in the dataframe.

Lets also include `bmi` in the plot:
```{r}
data %>% ggplot(aes(x = as.numeric(rownames(.)))) +
        geom_point(aes(y = weight_first)) +
        geom_point(aes(y = bmi, col = "red")) +
        theme_classic()
```
Here we see that the missing chunk for the is at the end of the data frame.
It starts at the row number 1387.

Does this coincide with any patterns in other variables?

Lets calculate means for all the numeric variables, for the part of the data, where
the `bmi` values are missing and the rest and then compare them:
```{r}
mean_missing <- data %>% select_if(is.numeric) %>% 
        filter(as.numeric(rownames(.)) >= 1387) %>% 
        summarise(colname = colnames(.), mean = colMeans(., na.rm = TRUE)) %>% 
        mutate(in_chunk = TRUE)
        

mean_missing <- data %>% select_if(is.numeric) %>% 
        filter(as.numeric(rownames(.)) < 1387) %>% 
        summarise(colname = colnames(.), mean = colMeans(., na.rm = TRUE)) %>% 
        mutate(in_chunk = FALSE) %>% 
        rbind(mean_missing)
        
mean_missing %>% ggplot(aes(x = colname, y = mean, fill = as.factor(in_chunk))) +
        geom_bar(stat = "identity", position = "dodge") +
        theme_classic() +
        coord_flip()
```
There arent any big differences in the means between the part of the dataframe, where `bmi`
is missing and the rest.

Lets look at the factor variables:
```{r}
mean_missing_f <- data %>% select_if(is.factor) %>% 
        sapply(.,as.numeric) %>% 
        as.data.frame() %>% 
        filter(as.numeric(rownames(.)) >= 1387) %>% 
        summarise(colname = colnames(.), mean = colMeans(., na.rm = TRUE)) %>% 
        mutate(in_chunk = TRUE)

mean_missing_f <- data %>% select_if(is.factor) %>% 
        sapply(.,as.numeric) %>% 
        as.data.frame() %>%
        filter(as.numeric(rownames(.)) < 1387) %>% 
        summarise(colname = colnames(.), mean = colMeans(., na.rm = TRUE)) %>% 
        mutate(in_chunk = FALSE) %>% 
        rbind(mean_missing_f)

mean_missing_f %>% ggplot(aes(x = colname, y = mean, fill = as.factor(in_chunk))) +
        geom_bar(stat = "identity", position = "dodge") +
        theme_classic() +
        coord_flip()
```
Here we just converted the factor levels into numeric values and then took the means.
This is only a quick solution. For factors with mor ethen two levels, this doesnt 
necessarily show the differences perfectly.

It  seems here as well there are no obvious patterns in the other variables.

The reason why exactly these values are missing remains unclear.

Is there any correlation between `weight_first` and `bmi`?
```{r}
data %>% ggplot(aes(x = weight_first, y = bmi)) +
        geom_point() +
        theme_classic()
```


```{r}
cor.test(data$bmi, data$weight_first)
```

-------

The main question we will pose is: Which of the available variables in the data set are useful
for predicting death within 28 days, after the use of an indwelling arterial catheter (IAC)?

It is important to note that these risk factors will likely not be very specific
for the use of AIC but might be just general risk factors.

Are there sufficient records for patients on which an IAC was used?
```{r}
table(data$aline_flg)
```

Subset the data and drop the `aline_flg` variable:
```{r}
data_sub <- data %>% filter(aline_flg == 1) %>% 
        select(!aline_flg)
```

Lets look at missing values.
Produce a missingnes plot for all the columns where there are missing values:
```{r}
missmap(data_sub[,colSums(is.na(data_sub)) > 0], main = "Missingness Map: Patients with AIC")
```



We will try to impute the missing values by using the mice::mice function.

It might be useful to have more data to do this, so we will do the imputation on
the whole data set.
Lets first look at a missingnes map of the whole dataset.


Fo the complete dataset, there is overall a bit more data missing and some variables
contain relatively more missing values. The overall patern (a chunk of missing values
for `bmi`) however, is the same.

Perform imputation only for subet data:
```{r message=FALSE, cache=TRUE}
#data_imp <- mice(data_sub)
```

```{r}
#plot(data_imp)
```


- Use vif to assess multicolinearity.